# Demography Informal Methods Seminar - Classification and Regression Trees (tidymodels)

```{r}
#| include: false
knitr::opts_chunk$set(echo = TRUE)
```

## Classification models

I would suggest you read section 5.1 of [Introduction to Statistical Learning](https://utsacloud-my.sharepoint.com/:b:/g/personal/corey_sparks_utsa_edu/EcwJ43xnPG5Jg2XquDnHdmkBPstgvNIyqTTugCfEeFGDFQ?e=SSlWyT) to get a full treatment of this topic

In classification methods, we are typically interested in using some observed characteristics of a case to predict a binary categorical outcome. This can be extended to a multi-category outcome, but the largest number of applications involve a 1/0 outcome.

In these examples, we will use the Demographic and Health Survey [Model Data](https://dhsprogram.com/data/Model-Datasets.cfm). These are based on the DHS survey, but are publicly available and are used to practice using the DHS data sets, but don't represent a real country.

In this example, we will use the outcome of contraceptive choice (modern vs other/none) as our outcome.

```{r}
library(haven)
dat <- url("https://github.com/coreysparks/data/blob/master/ZZIR62FL.DTA?raw=true")
model.dat <- read_dta(dat)
```

Here we recode some of our variables and limit our data to those women who are not currently pregnant and who are sexually active.

```{r}
library(dplyr)

model.dat2 <- model.dat %>%
  mutate(
    region      = v024,
    modcontra   = ifelse(v364 == 1, 1, 0),
    age         = cut(v012, breaks = 5),
    livchildren = v218,
    educ        = v106,
    currpreg    = v213,
    wealth      = as.factor(v190),
    partnered   = ifelse(v701 <= 1, 0, 1),
    work        = ifelse(v731 %in% c(0, 1), 0, 1),
    knowmodern  = ifelse(v301 == 3, 1, 0),
    age2        = v012^2,
    rural       = ifelse(v025 == 2, 1, 0),
    wantmore    = ifelse(v605 %in% c(1, 2), 1, 0)
  ) %>%
  filter(currpreg == 0, v536 > 0, v701 != 9) %>%
  dplyr::select(caseid, region, modcontra, age, age2, livchildren, educ,
                knowmodern, rural, wantmore, partnered, wealth, work)
```

```{r}
#| results: asis
knitr::kable(head(model.dat2))
```


## Cross-validation of predictive models

The term [cross-validation](https://www.cs.cmu.edu/~schneide/tut5/node42.html) refers to fitting a model on a subset of data and then testing it on another subset of the data. Typically this process is repeated several times.

**K-fold** cross-validation is a process where you leave out a "group" of observations, it is as follows:

1. Randomize the data
2. Split the data into k groups, where k is an integer
3. For each of the k groups,
    + Take one of the groups as a hold out test set
    + Use the other k-1 groups as training data
    + Fit a model using the data on the k-1 groups, and test it on the hold out group
    + Measure predictive accuracy of that model, and **throw the model away!**
4. Summarize the model accuracy over the measured model accuracy metrics


## Alternative accuracy measures

The Receiver Operating Characteristics (**ROC**) curve is a probability curve commonly used with the area under the curve (**AUC**) measure to summarize the separability of the classes. The higher the AUC, the better the model is at distinguishing between classes.


## Simple regression tree example

### Hitters data

```{r}
#| fig-width: 7
#| fig-height: 5
#| message: false
#| warning: false
library(tidymodels)
library(rpart.plot)
data(Hitters, package = "ISLR")

# Fit a single rpart tree for visualization
hitters_clean <- Hitters %>% filter(!is.na(Salary))

hitters_rec <- recipe(Salary ~ Years + Hits, data = hitters_clean)

hitters_spec <- decision_tree(cost_complexity = tune(), tree_depth = tune()) %>%
  set_engine("rpart") %>%
  set_mode("regression")

hitters_wf <- workflow(hitters_rec, hitters_spec)

# Quick visualization with a fixed tree (depth = 3)
hitters_vis_spec <- decision_tree(tree_depth = 3) %>%
  set_engine("rpart") %>%
  set_mode("regression")

hitters_vis_fit <- workflow(hitters_rec, hitters_vis_spec) %>%
  fit(hitters_clean)

prp(extract_fit_engine(hitters_vis_fit), type = 4, extra = 1,
    main = "Regression tree: Salary ~ Years + Hits")
```

Cross-validate to find the best cost-complexity (tree size):

```{r}
#| fig-width: 6
#| fig-height: 4
#| warning: false
set.seed(1115)
hitters_folds <- vfold_cv(hitters_clean, v = 10)

hitters_grid <- grid_regular(
  cost_complexity(range = c(-4, -1)),
  tree_depth(range = c(1, 6)),
  levels = 5
)

hitters_res <- tune_grid(hitters_wf, resamples = hitters_folds,
                         grid = hitters_grid, metrics = metric_set(rmse))

autoplot(hitters_res)
```

Prune to best tree and display:

```{r}
#| fig-width: 7
#| fig-height: 5
best_hitters <- select_best(hitters_res, metric = "rmse")

hitters_final <- finalize_workflow(hitters_wf, best_hitters) %>%
  fit(hitters_clean)

prp(extract_fit_engine(hitters_final), type = 4, extra = 1,
    main = "Pruned regression tree")
```

Single prediction example — player with 6 years experience and 200 hits:

```{r}
new <- data.frame(Hits = 200, Years = 6)
predict(hitters_final, new_data = new)
```


## PRB data — regression tree and random forest

```{r}
#| echo: false
#| results: asis
#| message: false
library(dplyr)
load(url("https://github.com/coreysparks/r_courses/raw/refs/heads/master/data/prb_2008.Rdata"))
names(prb) <- tolower(names(prb))
```

```{r}
#| results: hide
#| message: false
#| echo: false
library(mice)
prb2 <- prb %>%
  dplyr::select(-y, -x, -id, -country, -region, -year, -e0male, -e0female,
                -popdenspersqmile, -percpop1549hivaids2001,
                -projectedpopmid2025, -projectedpopmid2050,
                -projectedpopchange_08_50perc)
imps  <- mice(m = 10, data = prb2)
prb2  <- complete(imps)
```

### Simple regression tree — PRB data

```{r}
#| fig-width: 7
#| fig-height: 6
#| warning: false
#| message: false
set.seed(1115)
prb_split <- initial_split(prb2 %>% dplyr::select(-lowe0), prop = 0.75)
prb_train <- training(prb_split)
prb_test  <- testing(prb_split)

prb_rec  <- recipe(e0total ~ ., data = prb_train)

prb_tree_spec <- decision_tree(tree_depth = 7) %>%
  set_engine("rpart") %>%
  set_mode("regression")

prb_tree_fit <- workflow(prb_rec, prb_tree_spec) %>%
  fit(prb_train)

prp(extract_fit_engine(prb_tree_fit), type = 4, extra = 1,
    main = "Regression tree: e0total")
```

### Bagged regression tree — PRB data

```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
#| warning: false
library(baguette)

prb_bag_spec <- bag_tree(min_n = 5) %>%
  set_engine("rpart", times = 100) %>%
  set_mode("regression")

prb_bag_fit <- workflow(prb_rec, prb_bag_spec) %>%
  fit(prb_train)

prb_bag_fit
```

### Random forest — PRB data

```{r}
#| fig-width: 7
#| fig-height: 6
#| message: false
#| warning: false
library(ranger)

prb_rf_spec <- rand_forest(mtry = tune(), trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

prb_rf_wf <- workflow(prb_rec, prb_rf_spec)

set.seed(1115)
prb_folds <- vfold_cv(prb_train, v = 5)

prb_rf_res <- tune_grid(
  prb_rf_wf,
  resamples = prb_folds,
  grid      = grid_regular(mtry(range = c(2, 14)), levels = 7),
  metrics   = metric_set(rmse)
)

autoplot(prb_rf_res)
```

```{r}
#| fig-width: 7
#| fig-height: 6
best_prb_rf  <- select_best(prb_rf_res, metric = "rmse")
prb_rf_final <- finalize_workflow(prb_rf_wf, best_prb_rf) %>%
  fit(prb_train)

# Variable importance
library(vip)
vip(prb_rf_final, num_features = 10)
```

### Classification tree — low vs high life expectancy

```{r}
#| fig-width: 7
#| fig-height: 6
#| warning: false
#| message: false
prb2_cls <- prb2 %>%
  mutate(lowe0 = as.factor(ifelse(e0total < median(e0total), "low", "high")))

set.seed(1115)
prb_cls_split <- initial_split(prb2_cls %>% dplyr::select(-e0total), prop = 0.75)
prb_cls_train <- training(prb_cls_split)
prb_cls_test  <- testing(prb_cls_split)

prb_cls_rec <- recipe(lowe0 ~ ., data = prb_cls_train)

prb_cls_spec <- decision_tree(cost_complexity = tune()) %>%
  set_engine("rpart") %>%
  set_mode("classification")

prb_cls_wf <- workflow(prb_cls_rec, prb_cls_spec)

set.seed(1115)
prb_cls_folds <- vfold_cv(prb_cls_train, v = 10)

prb_cls_res <- tune_grid(
  prb_cls_wf,
  resamples = prb_cls_folds,
  grid      = grid_regular(cost_complexity(range = c(-4, -1)), levels = 10),
  metrics   = metric_set(accuracy, roc_auc)
)

best_cp <- select_best(prb_cls_res, metric = "accuracy")

prb_cls_final <- finalize_workflow(prb_cls_wf, best_cp) %>%
  fit(prb_cls_train)

prp(extract_fit_engine(prb_cls_final), type = 4, extra = 4,
    main = "Classification tree: low vs high life expectancy")
```

### Random forest classification — PRB low life expectancy

```{r}
#| fig-width: 7
#| fig-height: 6
prb_rf_cls_spec <- rand_forest(mtry = tune(), trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")

prb_rf_cls_wf <- workflow(prb_cls_rec, prb_rf_cls_spec)

set.seed(1115)
prb_rf_cls_res <- tune_grid(
  prb_rf_cls_wf,
  resamples = prb_cls_folds,
  grid      = grid_regular(mtry(range = c(2, 14)), levels = 7),
  metrics   = metric_set(accuracy, roc_auc)
)

best_prb_rf_cls  <- select_best(prb_rf_cls_res, metric = "roc_auc")
prb_rf_cls_final <- finalize_workflow(prb_rf_cls_wf, best_prb_rf_cls) %>%
  fit(prb_cls_train)

# Accuracy on test set
prb_cls_preds <- predict(prb_rf_cls_final, prb_cls_test) %>%
  bind_cols(prb_cls_test %>% dplyr::select(lowe0))

accuracy(prb_cls_preds, truth = lowe0, estimate = .pred_class)
conf_mat(prb_cls_preds, truth = lowe0, estimate = .pred_class)

vip(prb_rf_cls_final, num_features = 10)
```


## More complicated example — DHS contraception data

### Train / test split

We use an 80% training fraction.

```{r}
#| message: false
#| warning: false
library(tidymodels)
set.seed(1115)

model.dat2 <- model.dat2 %>%
  mutate(modcontra = as.factor(ifelse(modcontra == 1, "mod", "notmod")))

contra_split <- initial_split(model.dat2 %>% dplyr::select(-caseid),
                              prop = 0.80, strata = modcontra)
dtrain <- training(contra_split)
dtest  <- testing(contra_split)

prop.table(table(dtrain$modcontra))
prop.table(table(dtest$modcontra))
```

### Recipe — preprocessing

The recipe replaces the manual `model.matrix()` call from the original analysis. We one-hot encode factor predictors, then center and scale all numeric predictors.

```{r}
contra_rec <- recipe(modcontra ~ region + age + age2 + livchildren + educ +
                       knowmodern + rural + wantmore + partnered + wealth + work,
                     data = dtrain) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

contra_rec
```

### Resampling — repeated 5-fold CV with downsampling

```{r}
#| message: false
library(themis)

set.seed(1115)
contra_folds <- vfold_cv(dtrain, v = 5, repeats = 5, strata = modcontra)

# Add downsampling inside the recipe to handle class imbalance
contra_rec_ds <- contra_rec %>%
  step_downsample(modcontra, under_ratio = 1, seed = 1115)
```

### Model specifications

```{r}
# Classification tree (rpart)
ct_spec <- decision_tree(cost_complexity = tune()) %>%
  set_engine("rpart") %>%
  set_mode("classification")

# Bagged tree
bt_spec <- bag_tree(min_n = tune()) %>%
  set_engine("rpart", times = 50) %>%
  set_mode("classification")

# Random forest
rf_spec <- rand_forest(mtry = tune(), trees = 500) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("classification")
```

### Workflows

```{r}
ct_wf <- workflow(contra_rec_ds, ct_spec)
bt_wf <- workflow(contra_rec_ds, bt_spec)
rf_wf <- workflow(contra_rec_ds, rf_spec)
```

### Train classification tree (rp1 equivalent)

```{r}
#| fig-width: 8
#| fig-height: 6
#| message: false
#| warning: false
ct_grid <- grid_regular(cost_complexity(range = c(-4, -1)), levels = 5)

set.seed(1115)
rp1 <- tune_grid(
  ct_wf,
  resamples = contra_folds,
  grid      = ct_grid,
  metrics   = metric_set(roc_auc, accuracy)
)

autoplot(rp1)
```

```{r}
#| fig-width: 8
#| fig-height: 6
best_ct <- select_best(rp1, metric = "roc_auc")
ct_final <- finalize_workflow(ct_wf, best_ct) %>% fit(dtrain)

prp(extract_fit_engine(ct_final), type = 4, extra = 4,
    main = "Classification tree for using modern contraception")
```

```{r}
#| fig-width: 7
#| fig-height: 5
vip(ct_final, num_features = 10, geom = "col")
```

#### SHAP values for classification tree model

SHAP (SHapley Additive exPlanations) values show how much each predictor contributes to an individual prediction and in which direction.

```{r}
#| fig-width: 8
#| fig-height: 6
#| message: false
#| warning: false
library(fastshap)
library(shapviz)

x_train <- bake(prep(contra_rec), new_data = dtrain) %>%
  dplyr::select(-modcontra)

pfun_ct <- function(object, newdata) {
  predict(object, new_data = newdata, type = "prob")$.pred_mod
}

set.seed(1115)
shap_ct <- explain(
  object       = ct_final,
  X            = x_train,
  pred_wrapper = pfun_ct,
  nsim         = 50,
  adjust       = TRUE
)

sv_ct <- shapviz(shap_ct, X = x_train)
sv_importance(sv_ct, kind = "beeswarm", max_display = 15)
```

```{r}
sv_importance(sv_ct, kind = "bar", max_display = 15) +
  labs(title = "Mean |SHAP| — classification tree, top 15 features")
```

### Train/evaluate each model — training set metrics

```{r}
#| message: false
# Helper that computes the same metrics as the original classify_metrics()
tidy_metrics <- function(preds_tbl, truth_col = "modcontra") {
  truth_sym <- sym(truth_col)
  bind_rows(
    accuracy(preds_tbl,   truth = !!truth_sym, estimate = .pred_class),
    sens(preds_tbl,       truth = !!truth_sym, estimate = .pred_class, event_level = "first"),
    spec(preds_tbl,       truth = !!truth_sym, estimate = .pred_class, event_level = "first"),
    precision(preds_tbl,  truth = !!truth_sym, estimate = .pred_class, event_level = "first"),
    f_meas(preds_tbl,     truth = !!truth_sym, estimate = .pred_class, event_level = "first")
  )
}
```

```{r}
##  Classification tree — train accuracy
ct_train_preds <- predict(ct_final, dtrain) %>%
  bind_cols(dtrain %>% dplyr::select(modcontra))

conf_mat(ct_train_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(ct_train_preds)
```

```{r}
## Classification tree — test accuracy
ct_test_preds <- predict(ct_final, dtest) %>%
  bind_cols(dtest %>% dplyr::select(modcontra))

conf_mat(ct_test_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(ct_test_preds)
```

### Train bagged tree model

```{r}
#| message: false
#| warning: false
bt_grid <- data.frame(min_n = c(2, 5, 10, 20))

set.seed(1115)
bt_res <- tune_grid(
  bt_wf,
  resamples = contra_folds,
  grid      = bt_grid,
  metrics   = metric_set(roc_auc, accuracy)
)

best_bt <- select_best(bt_res, metric = "roc_auc")
bt_final <- finalize_workflow(bt_wf, best_bt) %>% fit(dtrain)

bt_final
vip(bt_final, num_features = 10, geom = "col")
```

```{r}
## Bagged tree — train accuracy
bt_train_preds <- predict(bt_final, dtrain) %>%
  bind_cols(dtrain %>% dplyr::select(modcontra))

conf_mat(bt_train_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(bt_train_preds)

## Bagged tree — test accuracy
bt_test_preds <- predict(bt_final, dtest) %>%
  bind_cols(dtest %>% dplyr::select(modcontra))

conf_mat(bt_test_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(bt_test_preds)
```

### Train random forest model

```{r}
#| message: false
#| warning: false
set.seed(1115)
rf_res <- tune_grid(
  rf_wf,
  resamples = contra_folds,
  grid      = grid_regular(mtry(range = c(2, 20)), levels = 7),
  metrics   = metric_set(roc_auc, accuracy)
)

autoplot(rf_res)
```

```{r}
best_rf  <- select_best(rf_res, metric = "roc_auc")
rf_final <- finalize_workflow(rf_wf, best_rf) %>% fit(dtrain)

rf_final

vip(rf_final, num_features = 10, geom = "col")
```

```{r}
## Random forest — train accuracy
rf_train_preds <- predict(rf_final, dtrain) %>%
  bind_cols(dtrain %>% dplyr::select(modcontra))

conf_mat(rf_train_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(rf_train_preds)

## Random forest — test accuracy
rf_test_preds <- predict(rf_final, dtest) %>%
  bind_cols(dtest %>% dplyr::select(modcontra))

conf_mat(rf_test_preds, truth = modcontra, estimate = .pred_class)
tidy_metrics(rf_test_preds)
```

We see that by down-sampling the more common level of the outcome, we end up with much more balanced accuracy in terms of specificity and sensitivity.

### ROC curve

```{r}
#| fig-width: 7
#| fig-height: 5
library(pROC)

rf_test_prob <- predict(rf_final, dtest, type = "prob") %>%
  bind_cols(dtest %>% dplyr::select(modcontra))

roc_obj <- roc(
  response  = rf_test_prob$modcontra,
  predictor = rf_test_prob$.pred_mod,
  levels    = c("notmod", "mod")
)

plot.roc(roc_obj, grid = TRUE,
         main = paste0("ROC — random forest  (AUC = ", round(auc(roc_obj), 3), ")"))

auc(roc_obj)
```

### SHAP values for random forest model

```{r}
#| fig-width: 8
#| fig-height: 6
#| message: false
#| warning: false
pfun_rf <- function(object, newdata) {
  predict(object, new_data = newdata, type = "prob")$.pred_mod
}

set.seed(1115)
shap_rf <- explain(
  object       = rf_final,
  X            = x_train,
  pred_wrapper = pfun_rf,
  nsim         = 50,
  adjust       = TRUE
)

sv_rf <- shapviz(shap_rf, X = x_train)
sv_importance(sv_rf, kind = "beeswarm", max_display = 15)
```

```{r}
sv_importance(sv_rf, kind = "bar", max_display = 15) +
  labs(title = "Mean |SHAP| — random forest, top 15 features")
```
